An Efficient Minibatch Acceptance Test for Metropolis-Hastings

### 2016-10-12 aistats2017.tex
About 8.05 pages. Very close ...

With this approach, there is no guarantee that $u^* \geq 0$. However, we have
some flexibility in the choice of $\sigma$ in Equation~(\ref{eq:overall_corr_problem}).
As we decrease the variance of $X_{\rm norm}$, the variance of $X_{\rm corr}$
grows by the same amount and is in fact the result of convolution with a
Gaussian whose variance is the difference.  Thus as $\sigma$ decreases,
$C_\sigma(X)$ grows and approaches the derivative of a logistic function at
$\sigma = 0$. It retains some very weak negative values for $\sigma > 0$ but
removal of those values leads to very small error. Table~\ref{tab:xcorr} shows
that the errors between $X_{\rm norm}+X_{\rm corr}$ and $X_{\rm log}$ can be
made very small, approaching single floating precision (about $10^{-7}$), and
Algorithm~\ref{alg:our_algorithm} describes our procedure. A few points:

@@

We now derive error bounds for our M-H test and the approximate target
distribution it generates. (We treat $X_{\rm corr}$ samples from the exact
correction distribution based on the results in Table~\ref{tab:xcorr}.)

In the most similar prior works,~\citet{cutting_mh_2014} uses asymptotic
arguments based on the CLT to argue that its approximate acceptance test error
tends to zero as batch size increases, but no quantitative bounds are given.
In~\citet{icml2014c1_bardenet14}, explicit bounds are given, but they depend on
bounding:
\begin{equation}\label{eq:bad_bound}
    C_{\theta, \theta'} = \max_{1\leq i\leq N}|\log p(x_i\mid\theta') - \log p(x_i\mid\theta)|.
\end{equation}
It is possible to analytically determine $C_{\theta,\theta'}$ for logistic
regression, but it is unclear how to do so for more complicated models, and one
generally needs to use all $p(x_i\mid \theta')$ terms\footnote{The sample code
provided by the authors of~\citet{icml2014c1_bardenet14} in fact computes
$C_{\theta,\theta'}$ explicitly for each sample generated, i.e. it traverses the
entire data, thus providing no performance advantage over the complete test.} to
get $C_{\theta,\theta'}$.  In contrast, we use quantitative forms of the Central
Limit Theorem which rely on measurable statistics from a single minibatch. Thus
a sampler using our approach does not need to see data beyond the current
minibatch.

@@

\begin{lemma}\label{lem:quant_clt}
Let $X_1,\ldots,X_n$ be a set of zero-mean, independent, identically-distributed
random variables with sample mean $\bar{X}$ and sample variance $s^2_X$
where:
\begin{equation}
    \bar{X} = \frac{1}{n}\sum_{i=1}^nX_i, \quad s_X = \frac{1}{n}\left(\sum_{i=1}^n(X_i-\bar{X}^2\right)^{\frac{1}{2}}.
\end{equation}
This implies $t=\bar{X}/s_X$ has an approximate Student's distribution
which approaches a normal distribution in the limit. Then
\begin{equation}\label{eq:clt-bounds}
    \sup_x|{\rm Pr}(t<x) - \Phi(x)| \leq \frac{6.4E|X|^3+2E|X|}{\sqrt{n}}.
\end{equation}
\end{lemma}

@@

To apply this to our test, let $X_i = \Lambda_i(\theta,\theta') -
\Lambda(\theta,\theta')$, so that the $X_i$ are zero-mean, i.i.d. variables. If
instead of all $n$ samples, we only extract a subset of $b$ samples
corresponding to our minibatch, we can connect $\bar{X}$ with our $\Delta^*$
term:
\begin{equation}\label{eq:x-hat}
    \bar{X} = \Delta^*(\theta,\theta') - \Delta(\theta,\theta'),
\end{equation}
so that $s_X = s_{\Delta^*}$. This results in the following:

\begin{corollary}\label{cor:our_bound_delta_prime}
We can now substitute into Equation~(\ref{eq:clt-bounds2}) and displace by the mean, giving:
\begin{align*}
    \sup_y\left|{\rm Pr}(\Delta^*< y) - \Phi\left(\frac{y-\Delta}{s_{\Delta^*}}\right)\right| &\leq \frac{6.4E|X|^3+2E|X|}{\sqrt{b}} \\
    &= \epsilon(\theta,\theta',b).
\end{align*}
\end{corollary} %\vspace{-10pt}
Corollary~\ref{cor:our_bound_delta_prime} shows that the distribution of
$\Delta^*$ approximates a Normal distribution with mean $\Delta$ and variance
$s^2_{\Delta^*}$. Furthermore, it bounds the error with \emph{estimable
quantities}: both $E|X|$ and $E|X|^3$ can be estimated as means of $|\Lambda_i
- \Lambda|$ and $|\Lambda_i - \Lambda|^3$, respectively, on each minibatch. We
expect this will often be accurate enough on minibatches with hundreds or
thousands of points, but otherwise bootstrap CIs can be computed from those
sequences.

% Daniel: I removed this sentence to save space.
% Since the bounds are monotone in $E|X|$ and $E|X|^3$, using upper
% bootstrap CI limits will provide high-confidence error bounds.

\subsection{Error Bounds are Preserved After Adding Random Variables}\label{ssec:preserve_bounds}

@@

\begin{lemma}\label{lem:cdf_bounds}
Let $P(x)$ and $Q(x)$ be two CDFs satisfying
$\sup_x|P(x)-Q(x)|\leq \epsilon$ with $x$ in some real range. Let $R(y)$ be the
{\em density} of another random variable $y$. Let $P'$ be the convolution $P*R$
and $Q'$ be the convolution $Q*R$. Then $P'(z)$ (resp. $Q'(z)$) is the CDF of
sum $z=x+y$ of independent random variables $x$ with CDF $P(x)$ (resp. $Q(x)$)
and y with density $R(y)$.  Then
\begin{equation}
    \sup_x|P'(x)-Q'(x)|\leq \epsilon
\end{equation}
\end{lemma}

\begin{proof}
See Supplementary Material, Section~\ref{app:proof_cdf_bounds}.
\end{proof}

@@

From Section~\ref{sec:our_algorithm}, as the distribution of $\Delta^*$
approaches a Gaussian, our new MH test becomes more accurate.
Corollary~\ref{cor:bounds_preserved} shows that the bounds from
Section~\ref{ssec:delta_star_distribution} are preserved after the addition of
the random variables we use, showing that our test should remain accurate.

% Daniel: I'm moving this paragraph over here and putting the rest of it in the
% supplementary material since our code doesn't do this, so we should not
% describe this in much detail.
In fact we can do better ($O(n^{-1})$) by using a more precise limit
distribution under an additional assumption. We review this in the Supplementary
Material, Section~\ref{app:better_error_bound}, and leave the details to future
work.

@@

We conduct two sets of experiments. In Section~\ref{ssec:gaussians}, we analyze
convergence to the posterior of a heated Gaussian mixture. In
Section~\ref{ssec:logistic}, we analyze its efficiency on logistic regression.

@@

We run MCMC with our M-H test and benchmark with~\citet{cutting_mh_2014}
and~\citet{icml2014c1_bardenet14}, all with initial minibatch size $m=100$.  For
the former, we increment minibatches by 100 and set the tolerance
$\epsilon=0.005$. For the latter, we increase sizes geometrically with a ratio
of $\gamma = 1.5$ and use error parameters $p = 2$ and $\delta = 0.01$.  All
methods collect 5000 samples using the same random walk proposer with covariance
${\rm diag}(0.15, 0.15)$, which means all shaping of distribution of the samples
is due to the M-H test.

@@

We discretize the posterior coordinates into bins with respect to the two
components of $\theta$.  The probability $P_i$ of a sample falling into bin $i$
is the integral of the true posterior probability over the area of that bin.  A
single sample from any of the MH methods should therefore be multinomial with
distribution $P$, and a set of $n$ (ideally independent) samples should be ${\rm
Multinomial}(P,n)$.  The ideal distribution is simple, so we can use it to
measure the quality of the sample distributions rather than use general purpose
tests like KL-divergence or likelihood-ratio, which can be problematic with zero
counts in some bins as we have here.

For large $n$, the per-bin distributions are approximated by Poissons with
parameter $\lambda_i=P_i n$. Given samples $\{\theta_1,\ldots,\theta_T\}$, let
$c_j$ denote the number of individual samples $\theta_i$ that fall in bin $j$
out of $N_{\rm bins}$ total. We have
\begin{equation}\label{eq:log_prob_poissons}
\begin{split}
    \log p(c_1, \ldots, c_{N_{\rm bins}} \mid P_1, \ldots, P_{N_{\rm bins}}) & =  \\
    \sum_{j=1}^{N_{\rm bins}} c_j \log (n P_j) - n P_j - \log(\Gamma(c_j+1)). & \\
\end{split}
\end{equation}
Table~\ref{tab:poissons} shows the results. It is difficult, however, to
interpret the scores, so we perform significance tests to show the difference
between the MCMC-sampled distributions and ground-truth using the Chi-Squared
distribution as the test statistic (also in Table~\ref{tab:poissons}). Both
imply that our method is slightly superior to~\citet{cutting_mh_2014}, but
slightly worse than~\citet{icml2014c1_bardenet14}.

Figure~\ref{fig:gauss_mix_2}, however, suggests that our method dominates in
terms of speed and efficiency. It shows histograms of the (final) minibatch
sizes used each iteration. Our method consumes significantly less data; most
sizes are smaller than 1000, and the average size is 210.  The other methods
occasionally need to consume nearly all data points, and average minibatch sizes
are 15562 and 16857. The average minibatch sizes roughly predict the running
times of these methods since all have a running time proportional to the total
data consumed, with the exception of~\citet{icml2014c1_bardenet14}, which
requires a pass over the data to compute $C_{\theta,\theta'}$ (see
Equation~(\ref{eq:bad_bound})). Using minibatch sizes as a proxy for time also
avoids discrepancies due to code optimization.

@@

We next use logistic regression for the binary classification of 1s versus 7s on
a subset of the MNIST8M dataset, which is a larger version of
MNIST~\citep{lecun-mnisthandwrittendigit-2010} and well-suited to our desire to
run MCMC on large data. We randomly subsampled 450k training and 192k testing
points. The pixels are scaled in $[0,1]$. We impose a flat prior on $\theta$ and
again use a random walk proposer, this time with covariance matrix $0.05I$ for
the $784\times 784$ identity matrix $I$. The posterior temperature is set at
$T=1000$.  We run our MH test for 3000 samples and again compare with the two
baseline algorithms with starting minibatch size 100. To
make~\citet{cutting_mh_2014} run faster, we use a larger $\epsilon=0.05$ value.
For~\citet{icml2014c1_bardenet14}, we tuned $\gamma \in \{1.1,1.25,1.5,1.75,2\}$
before concluding that $\gamma=1.5$ was reasonable. In addition, we originally
used their suggested analytic approximation to $C_{\theta,\theta'}$, but found
that the resulting values were too high, meaning that their method needed to
consume the entire dataset each iteration. Thus, we naively compute
$C_{\theta,\theta'}$.

@@

Figure~\ref{fig:logistic_minibatch}, in a similar manner as
Figure~\ref{fig:gauss_mix_2}, shows the histogram of minibatch sizes for all
three methods on a log-log scale. With an initial size of 100, our method
achieves an average minibatch size of 393, far smaller than the averages of the
other two methods, which is due to their longer-tailed distributions.

>>>>

With this approach, there is no guarantee that $u^* \geq 0$. However, we have
some flexibility in the choice of $\sigma$ in Equation~(\ref{eq:overall_corr_problem}).
As we decrease the variance of $X_{\rm norm}$, the variance of $X_{\rm corr}$
grows by the same amount and is in fact the result of convolution with a
Gaussian whose variance is the difference.  Thus as $\sigma$ decreases,
$C_\sigma(X)$ grows and approaches the derivative of a logistic function at
$\sigma = 0$. It retains some weak negative values for $\sigma > 0$ but removal
of those leads to small error. Table~\ref{tab:xcorr} shows that the errors
between $X_{\rm norm}+X_{\rm corr}$ and $X_{\rm log}$ approach single floating
precision (about $10^{-7}$), and Algorithm~\ref{alg:our_algorithm} describes our
procedure. A few points:

@@

We now derive error bounds for our M-H test and the approximate target
distribution it generates. In the most similar prior
works,~\citet{cutting_mh_2014} uses asymptotic arguments based on the CLT to
argue that its approximate acceptance test error tends to zero as batch size
increases, but no quantitative bounds are given.
In~\citet{icml2014c1_bardenet14}, explicit bounds are given, but they depend on
bounding:
\begin{equation}\label{eq:bad_bound}
    C_{\theta, \theta'} = \max_{1\leq i\leq N}|\log p(x_i\mid\theta') - \log p(x_i\mid\theta)|.
\end{equation}
It is possible to analytically determine $C_{\theta,\theta'}$ for logistic
regression, but it is unclear how to do so for more complicated models, and one
generally needs to use all $p(x_i\mid \theta')$ terms\footnote{The sample code
provided by the authors of~\citet{icml2014c1_bardenet14} in fact computes
$C_{\theta,\theta'}$ explicitly for each sample generated, i.e. it traverses the
entire data, thus providing no performance advantage over the complete test.} to
get $C_{\theta,\theta'}$.  In contrast, we use quantitative forms of the Central
Limit Theorem which rely on measurable statistics from a single minibatch. Thus
a sampler using our approach does not need to see data beyond the current
minibatch.

@@

\begin{lemma}\label{lem:quant_clt}
Let $X_1,\ldots,X_n$ be a set of zero-mean, independent, identically-distributed
random variables with sample mean $\bar{X}$ and sample variance $s^2_X$
where:
\begin{equation}
    \bar{X} = \frac{1}{n}\sum_{i=1}^nX_i, \quad s_X = \frac{1}{n}\left(\sum_{i=1}^n(X_i-\bar{X})^2\right)^{\frac{1}{2}}.
\end{equation}
This implies $t=\bar{X}/s_X$ has an approximate Student's distribution
which approaches a normal distribution in the limit. Then
\begin{equation}\label{eq:clt-bounds}
    \sup_x|{\rm Pr}(t<x) - \Phi(x)| \leq \frac{6.4E|X|^3+2E|X|}{\sqrt{n}}.
\end{equation}
\end{lemma}

@@

To apply this to our test, let $X_i = \Lambda_i(\theta,\theta') -
\Lambda(\theta,\theta')$, so that the $X_i$ are zero-mean, i.i.d. variables. If
instead of all $n$ samples, we only extract a subset of $b$ samples
corresponding to our minibatch, we can connect $\bar{X}$ with our $\Delta^*$
term: $\bar{X} = \Delta^*(\theta,\theta') - \Delta(\theta,\theta')$, so that
$s_X = s_{\Delta^*}$. This results in the following:

\begin{corollary}\label{cor:our_bound_delta_prime}
We can now substitute into Equation~(\ref{eq:clt-bounds2}) and displace by the mean, giving:
\begin{align*}
    \sup_y\left|{\rm Pr}(\Delta^*< y) - \Phi\left(\frac{y-\Delta}{s_{\Delta^*}}\right)\right| &\leq \frac{6.4E|X|^3+2E|X|}{\sqrt{b}} \\
    &= \epsilon(\theta,\theta',b).
\end{align*}
\end{corollary} %\vspace{-10pt}
Corollary~\ref{cor:our_bound_delta_prime} shows that the distribution of
$\Delta^*$ approximates a Normal distribution with mean $\Delta$ and variance
$s^2_{\Delta^*}$. Furthermore, it bounds the error with \emph{estimable
quantities}: both $E|X|$ and $E|X|^3$ can be estimated as means of $|\Lambda_i
- \Lambda|$ and $|\Lambda_i - \Lambda|^3$, respectively, on each minibatch. We
expect this will often be accurate enough on minibatches with hundreds of
points, but otherwise bootstrap CIs can be computed.

% Daniel: Lame, I know, but it saves one line.
%\subsection{Error Bounds are Preserved After Adding Random Variables}\label{ssec:preserve_bounds}
\subsection{Adding Random Variables}\label{ssec:preserve_bounds}

@@

\begin{lemma}\label{lem:cdf_bounds}
Let $P(x)$ and $Q(x)$ be two CDFs satisfying
$\sup_x|P(x)-Q(x)|\leq \epsilon$ with $x$ in some real range. Let $R(y)$ be the
{\em density} of another random variable $y$. Let $P'$ be the convolution $P*R$
and $Q'$ be the convolution $Q*R$. Then $P'(z)$ (resp. $Q'(z)$) is the CDF of
sum $z=x+y$ of independent random variables $x$ with CDF $P(x)$ (resp. $Q(x)$)
and y with density $R(y)$.  Then
\begin{equation}
    \sup_x|P'(x)-Q'(x)|\leq \epsilon.
\end{equation}
\end{lemma}
\begin{proof}
See Supplementary Material, Section~\ref{app:proof_cdf_bounds}.
\end{proof}

@@

From Section~\ref{sec:our_algorithm}, as the distribution of $\Delta^*$
approaches a Gaussian, our new MH test becomes more accurate.
Corollary~\ref{cor:bounds_preserved} shows that the bounds from
Section~\ref{ssec:delta_star_distribution} are preserved after the addition of
the random variables we use, showing that our test should remain accurate.
% Daniel: I'm moving this paragraph over here and putting the rest of it in the
% supplementary material since our code doesn't do this, so we should not
% describe this in much detail.
In fact we can do better ($O(n^{-1})$) by using a more precise limit
distribution under an additional assumption. We review this in the Supplementary
Material, Section~\ref{app:better_error_bound}, and leave the details to future
work.

@@

We conduct two sets of experiments analyzing convergence to the posterior of a
heated Gaussian mixture, and efficiency on logistic regression.

@@

We run MCMC with our test and benchmark with~\citet{cutting_mh_2014}
and~\citet{icml2014c1_bardenet14}, all with initial minibatch size 100.  For the
former, we increment sizes by 100 and set the tolerance $\epsilon=0.005$.  For
the latter, we increase sizes geometrically with $\gamma = 1.5$ and use
parameters $p = 2, \delta = 0.01$.  All methods collect 5000 samples using a
random walk proposer with covariance ${\rm diag}(0.15, 0.15)$, which means the
M-H test is responsible for the sample distribution.

@@

We discretize the posterior coordinates into bins with respect to the two
components of $\theta$.  The probability $P_i$ of a sample falling into bin $i$
is the integral of the true posterior over the bin's area.  A single sample
should therefore be multinomial with distribution $P$, and a set of $n$ (ideally
independent) samples is ${\rm Multinomial}(P,n)$. This distribution is simple
and we can use it to measure the quality of the samples rather than use general
purpose tests like KL-divergence or likelihood-ratio, which can be problematic
with zero counts in bins.

For large $n$, the per-bin distributions are approximated by Poissons with
parameter $\lambda_i=P_i n$. Given samples $\{\theta_1,\ldots,\theta_T\}$, let
$c_j$ denote the number of individual samples $\theta_i$ that fall in bin $j$
out of $N_{\rm bins}$ total. We have
\begin{equation}\label{eq:log_prob_poissons}
\begin{split}
    \log p(c_1, \ldots, c_{N_{\rm bins}} \mid P_1, \ldots, P_{N_{\rm bins}}) & =  \\
    \sum_{j=1}^{N_{\rm bins}} c_j \log (n P_j) - n P_j - \log(\Gamma(c_j+1)). & \\
\end{split}
\end{equation}
Table~\ref{tab:poissons} shows the results, but it is difficult to interpret the
scores, so we perform significance tests to show the difference between the
MCMC-samples and ground-truth using the Chi-Squared distribution as the test
statistic (also in Table~\ref{tab:poissons}). Both metrics imply that our method
is slightly superior to~\citet{cutting_mh_2014}, but slightly worse
than~\citet{icml2014c1_bardenet14}.

Figure~\ref{fig:gauss_mix_2}, however, suggests that our method dominates in
terms of speed and efficiency. The histograms of the (final) minibatch sizes
used each iteration show that our method consumes significantly less data; the
distribution is short-tailed and the mean is 210. The other methods occasionally
need to use nearly all data points, and average minibatch sizes are 15562 and
16857. The sizes roughly predict the running times of these methods, which are
proportional to the data consumed, with the exception
of~\citet{icml2014c1_bardenet14}, which requires a full pass to compute
$C_{\theta,\theta'}$ (Equation~(\ref{eq:bad_bound})). Using minibatch sizes
as a proxy for time also avoids discrepancies due to code optimization.

@@

We next use logistic regression for the binary classification of 1s versus 7s on
a subset of the MNIST8M dataset, a larger version of
MNIST~\citep{lecun-mnisthandwrittendigit-2010} and well-suited to our big data
objectives. We randomly subsampled 450k training and 192k testing points. The
pixels are scaled in $[0,1]$. We impose a flat prior on $\theta$ and again use a
random walk proposer, this time with covariance matrix $0.05I$ for the
$784\times 784$ identity matrix $I$. The temperature is set at $T=1000$. We run
the three tests for samples and again use 100 as the starting minibatch size. To
make~\citet{cutting_mh_2014} run faster, we now use $\epsilon=0.05$.
For~\citet{icml2014c1_bardenet14}, we tuned $\gamma \in \{1.1,1.25,1.5,1.75,2\}$
before concluding that $\gamma=2.0$ was OK. In addition, we tried their analytic
approximation to $C_{\theta,\theta'}$, but found that the bounds were too high,
meaning that their method needed to consume the entire dataset each iteration.
Thus, we naively compute $C_{\theta,\theta'}$.

@@

Figure~\ref{fig:logistic_minibatch} shows histograms of minibatch sizes for the
three methods on a log-log scale. With an initial size of 100, we achieve an
average minibatch size of 393, far smaller than the averages of the other two
methods, which have longer-tailed distributions.
